Supplement: Fremont Models

### 2023-11-28 R/analysis.qmd
minor changes to formatting, mostly added code to define function `prepare_image()`

---
title: "Supplement: Fremont Models"
date: today
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
execute: 
  eval: true
  echo: true
  warning: false
  error: false
---

@@

**Method**: a GAMM with a negative binomial distribution for count data with dispersion, an an offset to account for the area of each watershed, and an exponential covariance matrix to account for residual spatial autocorrelation.

@@

library(GGally)
library(ggeffects)
library(gt)
library(here)
library(mgcv)
library(patchwork)
library(sf)
library(sfdep)
library(tidyverse)
library(viridis)

@@

Dependent variable:  
- a count of **sites** per watershed

Independent variables:  
- **elevation**  
- maize growing degree days (**gdd**, in Celsius)  
- **precipitation** (millimeters)  
- cost-distance (in hours) to **springs** and **streams**  
- **protected** status  
- cost-distance (in hours) to **roads**

@@

Offsets:  
- **area** of each watershed

@@

utah <- read_sf(gpkg, "utah")

watersheds <- read_sf(gpkg, "watersheds") |> 
  mutate(
    huc4 = str_sub(id, 1, 4),
    huc4 = factor(huc4)
  )

@@

```{r}
#| fig-height: 4
#| code-fold: true
#| fig-align: center

max_density_obs <- with(watersheds, max(sites/area))

obs <- ggplot() +
  geom_sf(
    data = utah,
    fill = "gray95"
  ) +
  geom_sf(
    data = watersheds, 
    aes(fill = (sites/area)/max_density_obs),
    color = "gray90", 
    linewidth = 0.2
  ) +
  scale_fill_viridis(
    name = "Relative\nSite Density",
    option = "mako",
    limits = c(0, 1)
  ) +
  # coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.background = element_rect(fill = "gray95", color = "transparent"),
    legend.justification = c("left", "top"),
    legend.position = c(0.66, 0.57)
  )

obs

This map shows the relative density of sites based on their observed frequency in the study area. If site density is the ratio of the number of sites in a watershed to the area of that watershed, the _relative_ density is the density estimate for each watershed divided by the maximum density estimate across all watersheds. It is more appropriate to visualize this value as it doesn't have the same strong implications as visualizing the counts. (1) By visualizing the _density_, it does not make assumptions about the absence of sampling bias owing to the area of each watershed. (2) By visualizing the _relative_ density, it does not make assumptions about the total occurrence rate of sites across the study area, i,e, it does not include an estimate of the intercept.

@@

```{r}
#| fig-height: 5
#| fig-asp: 1
#| code-fold: true
#| out-width: "75%"
#| fig-align: center

@@

```{r}
#| fig-width: 8
#| fig-asp: 0.9
#| code-fold: true
#| fig-align: center

make_map <- function(fill, name) {

  ggplot() +
    geom_sf(
      data = utah,
      fill = "gray95"
    ) +
    geom_sf(
      data = watersheds, 
      aes(fill = {{fill}}),
      color = "gray90", 
      linewidth = 0.2
    ) +
    scale_fill_viridis(
      name = name,
      option = "mako"
    ) +
    theme_void() +
    theme(
      legend.background = element_rect(fill = "gray95", color = "transparent"),
      legend.justification = c("left", "bottom"),
      legend.position = c(0.65, 0.19)
    )

@@

for_print <- 
  (make_map(gdd, "GDD (°C)") + make_map(precipitation, "PPT (mm)")) /
  (make_map(streams, "Streams\n(hours)") + make_map(protected, "Protected"))

@@

ggsave(
  plot = for_print,
  filename = here("figures", "covariates.jpg"),
  width = 5.75,
  height = 5.75 * (dy/dx),
  dpi = 600
)

(make_map(gdd, "GDD (°C)") + 
  make_map(precipitation, "PPT (mm)") +
  make_map(streams, "Streams\n(hours)")) /
  (make_map(springs, "Springs\n(hours)") +
     make_map(protected, "Protected") +
     make_map(roads, "Roads\n(hours)"))

@@

```{r}
#| fig-width: 8
#| fig-asp: 0.45
#| code-fold: true
#| fig-align: center

@@

threshold_dists <- ggplot() +
  geom_histogram(
    data = values,
    aes(value),
    color = "white",
    binwidth = 100,
    center = 50
  ) +
  geom_vline(
    data = thresholds,
    aes(xintercept = xintercept),
    linewidth = 1.2,
    color = "darkred"
  ) +
  geom_text(
    data = thresholds,
    aes(x, y, label = label),
    color = "darkred",
    hjust = 0,
    vjust = 1
  ) +
  facet_wrap(
    ~variable, 
    scale = "free_x",
    labeller = labeller(
      variable = c(precipitation = "Precipitation (mm)", 
                   gdd = "Maize GDD (°C)")
    ),
    strip.position = "bottom"
  ) +
  scale_x_continuous(breaks = pretty_breaks) +
  labs(
    x = NULL,
    y = "Count of watersheds"
  ) +
  theme_bw(12) +
  theme(
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(size = rel(1)),
    strip.placement = "outside"
  )

ggsave(
  plot = threshold_dists,
  filename = here("figures", "threshold_distributions.jpg"),
  width = 5.75,
  height = 5.75 * (dy/(2*dx)),
  dpi = 600
)

threshold_dists

@@

```{r}
#| fig-width: 8
#| fig-asp: 0.45
#| code-fold: true
#| fig-align: center

threshold_map <- ggplot() +
    geom_sf(
      data = utah,
      fill = "gray95"
    ) +
    geom_sf(
      data = watersheds, 
      aes(fill = in_niche),
      color = "gray90", 
      linewidth = 0.2
    ) +
    scale_fill_viridis(
      name = "In Niche?",
      option = "mako",
      end = 0.75,
      discrete = TRUE
    ) +
    theme_void() +
    theme(
      legend.background = element_rect(fill = "gray95", color = "transparent"),
      legend.justification = c("left", "top"),
      legend.position = c(0.66, 0.57)
    )

obs + threshold_map

@@

```{r}
#| code-fold: true
#| fig-align: center
#| fig-height: 6
#| fig.asp: 0.93
#| out-width: "70%"

margins <- ggpredict(fremont) |>
  unclass() |>
  bind_rows() |>
  as_tibble() |> # the "ggeffects" class doesn't want to let go
  filter(group != "area") |>
  rename(
    "y" = predicted,
    "ymin" = conf.low,
    "ymax" = conf.high,
    "variable" = group
  ) |> 
  mutate(
    # hacky way to "clip" the ribbon before expanding in ggplot
    ymax = ifelse(ymax > 100, 100, ymax)
  )

@@

ggsave(
  plot = marginal_response,
  filename = here("figures", "marginal_response.jpg"),
  width = 5,
  height = 5 * 0.93,
  dpi = 600
)

@@

```{r}
#| fig-width: 8
#| fig-asp: 0.9
#| code-fold: true
#| fig-align: center

basemap <- ggplot() +
  geom_sf(
    data = utah,
    fill = "gray95"
  ) +
  theme_void() +
  theme(
    legend.background = element_rect(fill = "gray95", 
                                     color = "transparent"),
    legend.justification = c("left", "bottom"),
    legend.position = c(0.65, 0.19)
  )

obs <- basemap +
  geom_sf(
    data = watersheds, 
    aes(fill = (sites/area)/max_density_obs),
    color = "gray90", 
    linewidth = 0.1
  ) +
  scale_fill_viridis(
    name = "Observed",
    option = "mako",
    limits = c(0, 1)
  )

max_density_est <- with(watersheds, max(estimate/area))

estimate <- basemap +
  geom_sf(
    data = watersheds, 
    aes(fill = (estimate/area)/max_density_est),
    color = "gray90", 
    linewidth = 0.1
  ) +
  scale_fill_viridis(
    name = "Estimated",
    option = "mako",
    limits = c(0, 1)
  )

res <- basemap +
  geom_sf(
    data = watersheds, 
    aes(fill = residuals),
    color = "gray90", 
    linewidth = 0.1
  ) +
  scale_fill_viridis(
    name = "Residuals",
    option = "mako"
  )

ggsave(
  plot = (obs + estimate),
  filename = here("figures", "model-map.jpg"),
  width = 5.75,
  height = 5.75 * (dy/(2*dx)),
  dpi = 600
)

ggsave(
  plot = (estimate + threshold_map),
  filename = here("figures", "threshold_map.jpg"),
  width = 5.75,
  height = 5.75 * (dy/(2*dx)),
  dpi = 600
)

(obs + estimate) / (res + threshold_map)

@@

Please note that the residuals are still being shown for the count-level data.

>>>>

---
title: "Supplement: Fremont Models"
date: today
format:
  html: 
    embed-resources: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
execute: 
  eval: true
  echo: true
  warning: false
  error: false
---

```{css}
/*| echo: false */
figcaption {
    display: none !important;
}
```

@@

**Method**: a GAMM with a negative binomial distribution for count data with
dispersion, an offset to account for the area of each watershed, and an
exponential covariance matrix to account for residual spatial autocorrelation.

@@

library(GGally)
library(ggeffects)
library(gt)
library(here)
library(mgcv)
library(patchwork)
library(sf)
library(sfdep)
library(tidyverse)
library(tigris)
library(viridis)

Some plotting defaults

```{r}
#| code-fold: true
#| file: "fun-prepare-image.R"

@@

Dependent variable:\
- a count of **sites** per watershed

Independent variables:\
- **elevation**\
- maize growing degree days (**gdd**, in Celsius)\
- **precipitation** (millimeters)\
- cost-distance (in hours) to **springs** and **streams**\
- **protected** status\
- cost-distance (in hours) to **roads**

@@

Offsets:\
- **area** of each watershed

@@

watersheds <- read_sf(gpkg, "watersheds")

# the following shapes are just for visualization purposes
utah <- read_sf(gpkg, "utah")

window <- read_sf(gpkg, "window")

roads <- read_sf(gpkg, "roads") |> 
  filter(rttyp %in% c("U", "I")) |> 
  st_intersection(window)

towns <- tigris::places("Utah", progress_bar = FALSE) |> 
  rename_with(tolower) |> 
  select(name) |> 
  filter(name %in% c("Salt Lake City", "Cedar City", "Delta", "Richfield")) |> 
  arrange(name) |> 
  st_centroid()

@@

::: {#fig-overview}
![](../figures/overview.jpg){width="100%" fig-alt="overview map of project area"}

Overview map.
:::

@@

```{r}
#| label: fig-correlations
#| code-fold: true
#| out-width: "75%"
#| fig-height: 5
#| fig-asp: 1
#| fig-alt: "Pearson correlation matrix"
#| fig-cap: "Pearson correlation matrix"

@@

```{r}
#| label: fig-covariates
#| code-fold: true
#| fig-width: 8
#| fig-asp: 0.9
#| fig-alt: "Map of covariates"
#| fig-cap: "Map of covariates"

make_map <- function(x, fill, name) {

  ggplot() +
    geom_sf(
      data = utah,
      fill = "gray95"
    ) +
    geom_sf(
      data = x, 
      aes(fill = {{ fill }}),
      color = "gray90", 
      linewidth = 0.2
    ) +
    scale_fill_viridis(
      name = name,
      option = "mako"
    ) +
    geom_sf(
      data = roads |> filter(rttyp == "I"),
      color = "white",
      linewidth = 0.4
    ) +
    geom_sf(
      data = roads |> filter(rttyp == "U"),
      color = "white",
      linewidth = 0.2
    ) +
    geom_sf(
      data = towns,
      fill = "white",
      color = "gray35",
      shape = 21,
      size = 2
    ) +
    theme_void() +
    theme(
      legend.background = element_rect(fill = "gray95", color = "transparent"),
      legend.justification = c("left", "bottom"),
      legend.position = c(0.65, 0.19),
      legend.title = element_text(size = rel(0.9))
    )

@@

gdd <- make_map(
  watersheds,
  fill = gdd,
  name = "GDD (°C)"
)

ppt <- make_map(
  watersheds,
  fill = precipitation,
  name = "PPT (mm)"
)

streams <- make_map(
  watersheds,
  fill = streams,
  name = "Streams\n(hours)"
)

protected <- make_map(
  watersheds,
  fill = protected, 
  name = "Protected"
)

@@

fn <- here("figures", "covariates.jpg")

ggsave(
  plot = (gdd + ppt) / (streams + protected),
  filename = fn,
  width = 5.75,
  height = 5.75 * (dy/dx),
  dpi = 600
)

prepare_image(fn)

springs <- make_map(
  watersheds,
  fill = springs, 
  name = "Springs\n(hours)"
)

routes <- make_map(
  watersheds,
  fill = roads, 
  name = "Roads\n(hours)"
)

all_covariates <- (gdd + ppt + streams) / (springs + protected + routes)

remove(gdd, ppt, streams, springs, routes, protected)

all_covariates

@@

```{r}
#| label: fig-thresholds
#| code-fold: true
#| fig-width: 8
#| fig-asp: 0.45
#| fig-alt: "Histograms of temperature and precipitation"
#| fig-cap: "Histograms of temperature and precipitation"

@@

ggplot() +
  geom_histogram(
    data = values,
    aes(value),
    color = "white",
    binwidth = 100,
    center = 50
  ) +
  geom_vline(
    data = thresholds,
    aes(xintercept = xintercept),
    linewidth = 1.2,
    color = "darkred"
  ) +
  geom_text(
    data = thresholds,
    aes(x, y, label = label),
    color = "darkred",
    hjust = 0,
    vjust = 1
  ) +
  facet_wrap(
    ~variable, 
    scale = "free_x",
    labeller = labeller(
      variable = c(precipitation = "Precipitation (mm)", 
                   gdd = "Maize GDD (°C)")
    ),
    strip.position = "bottom"
  ) +
  scale_x_continuous(breaks = pretty_breaks) +
  labs(
    x = NULL,
    y = "Count of watersheds"
  ) +
  theme_bw(12) +
  theme(
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(size = rel(1)),
    strip.placement = "outside"
  )

@@

```{r}
#| label: fig-thresholds-map
#| code-fold: true
#| fig-width: 4
#| out-width: "4in"
#| fig-alt: "Map of maize niche thresholds"
#| fig-cap: "Map of maize niche thresholds"

thm <- ggplot() +
  geom_sf(
    data = utah,
    fill = "gray95"
  ) +
  geom_sf(
    data = watersheds, 
    aes(fill = in_niche),
    color = "gray90", 
    linewidth = 0.2
  ) +
  scale_fill_viridis(
    name = "In Niche?",
    option = "mako",
    end = 0.75,
    discrete = TRUE
  ) +
  geom_sf(
    data = roads |> filter(rttyp == "I"),
    color = "white",
    linewidth = 0.4
  ) +
  geom_sf(
    data = roads |> filter(rttyp == "U"),
    color = "white",
    linewidth = 0.2
  ) +
  geom_sf(
    data = towns,
    fill = "white",
    color = "gray35",
    shape = 21,
    size = 2
  ) +
  theme_void() +
  theme(
    legend.background = element_rect(fill = "gray95", color = "transparent"),
    legend.justification = c("left", "top"),
    legend.position = c(0.66, 0.57)
  )

thm

@@

margins <- ggpredict(fremont, condition = list(area = mean(watersheds$area))) |>
  unclass() |>
  bind_rows() |>
  as_tibble() |> # the "ggeffects" class doesn't want to let go
  filter(group != "area") |>
  rename(
    "y" = predicted,
    "ymin" = conf.low,
    "ymax" = conf.high,
    "variable" = group
  )

```{r}
#| label: fig-partials
#| code-fold: true
#| fig-height: 6
#| fig.asp: 0.93
#| out-width: "70%"
#| fig-alt: "Partial dependence plots"
#| fig-cap: "Partial dependence plots"

margins <- margins |> 
  mutate(
    # hacky way to "clip" the ribbon before expanding in ggplot
    ymax = ifelse(ymax > 100, 100, ymax)
  )

@@

fn <- here("figures", "marginal_response.jpg")

ggsave(
  plot = marginal_response,
  filename = fn,
  width = 5,
  height = 5 * 0.93,
  dpi = 600
)

prepare_image(fn)

@@

```{r}
#| label: fig-model-map
#| code-fold: true
#| fig-width: 8
#| fig-asp: 0.9
#| fig-alt: "Map of estimated density"
#| fig-cap: "Map of estimated density"

obs <- make_map(
  watersheds,
  fill = ifelse(sites == 0, log(0.00001), log(sites/area)), 
  name = "Observed\nLog Density"
)

est <- make_map(
  watersheds,
  log(estimate/area), 
  "Estimated\nLog Density"
)

res <- make_map(
  watersheds,
  fill = residuals, 
  name = "Normalized\nResiduals"
)

fn <- here("figures", "model-map.jpg")

ggsave(
  plot = (obs + est),
  filename = fn,
  width = 5.75,
  height = 5.75 * (dy/(2*dx)),
  dpi = 600
)

prepare_image(fn)

fn <- here("figures", "threshold_map.jpg")

ggsave(
  plot = (est + thm),
  filename = fn,
  width = 5.75,
  height = 5.75 * (dy/(2*dx)),
  dpi = 600
)

prepare_image(fn)

(obs + est) / (res + thm)

@@

Please note that these are normalized residuals from the LME model.
